---
title: "HSLS Analysis"
format: html
editor: visual
---

```{r}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(psych)
library(tidyr)

dathsls <- haven::read_sav("HSLS6.11.21.sav")

hsls <- dathsls

hsls <- hsls %>%
  rename(stu_id = STU_ID, # change column names
         sch_id = SCH_ID,
         # excellentTests = S1MTESTS,
         # understandTexts = S1MTEXTBOOK,
         # masterSkills = S1MSKILLS,
         # excellentAssign = S1MASSEXCL,
         sex = X1SEX,
         race = X1RACE,
         hispanic = X1HISPANIC,
         white = X1WHITE,
         black = X1BLACK,
         asian = X1ASIAN,
         pacificIsland = X1PACISLE,
         SES = X1SES,
         hsls_w_cohort_g9 = W1STUDENT,
         hsls_w_cohort_g12 = W2STUDENT)

```
# Give only the names that start with S1 or S2
S1 = 9th Grade 2009
S2 = 11th Grade 2012

```{r}

filtered_S1 <- names(hsls)[grep("^S1", names(hsls))]
filtered_S1

filtered_s2 <- names(hsls)[grep("^S2", names(hsls))]
filtered_s2




```

## Create subset of dataset with only math efficacy items
```{r}

names(hsls)[grep("^S1M|^S2M", names(hsls))]

math_eff <- hsls[, grep("^S1M|^S2M", names(hsls))]

# Ensures that the dataset was created successfully, difference should be 0
setdiff(names(hsls)[grep("^S1M | $S2M", names(hsls))], names(math_eff))

```
```{r}

# 1 Strongly agree 
# 2 Agree 
# 3 Disagree 
# 4 Strongly Disagree

ggplot(math_eff, aes(x = factor(S1MPERSON1))) +
  geom_bar(fill = "skyblue") +
  labs(title = "Distribution of S1MPERSON1 Responses",
       x = "S1MPERSON1",
       y = "Count") +
  theme_minimal()







```

```{r}

ggplot(hsls, aes(x = factor(S2MPERSON1))) +
  geom_bar(fill = "coral") +
  labs(title = "Distribution of S2MPERSON1 Responses",
       x = "S2MPERSON1",
       y = "Count") +
  theme_minimal()

```

# 
```{r}

long_SMPERSON1 <- gather(math_eff, key = "variable", value = "value", S1MPERSON1, S2MPERSON1)

# Create a grouped bar plot
ggplot(long_SMPERSON1, aes(x = factor(value), fill = variable)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of S1MPERSON1 and S2MPERSON1",
       x = "Likert Scale Response",
       y = "Count",
       fill = "Variable") +
  theme_minimal() +
  scale_fill_manual(values = c("skyblue", "coral")) +
  scale_x_discrete(labels = c("Strongly Agree", "Agree", "Disagree", "Strongly Disagree", "NA"))




```
```{r}


long_SMTESTS <- gather(math_eff, key = "variable", value = "value", S1MTESTS, S2MTESTS)

# Create a grouped bar plot
ggplot(long_SMTESTS, aes(x = factor(value), fill = variable)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of S1MTESTS and S2MTESTS",
       x = "Likert Scale Response",
       y = "Count",
       fill = "Variable") +
  theme_minimal() +
  scale_fill_manual(values = c("skyblue", "coral")) +
  scale_x_discrete(labels = c("Strongly Agree", "Agree", "Disagree", "Strongly Disagree", "NA"))







```
## Creating a function to automatically create graph of desired variables
```{r}

create_grouped_bar_plot <- function(data, missing_word) {
  # Dynamically gather the columns based on the missing word
  long_data <- gather(data, key = "variable", value = "value", 
                      !!paste0("S1M", missing_word), !!paste0("S2M", missing_word))
  
  # Create the grouped bar plot
  ggplot(long_data, aes(x = factor(value), fill = variable)) +
    geom_bar(position = "dodge") +
    labs(title = paste("Distribution of S1M", missing_word, " and S2M", missing_word, sep = ""),
         x = "Likert Scale Response",
         y = "Count",
         fill = "Variable") +
    theme_minimal() +
    scale_fill_manual(values = c("skyblue", "coral")) +
    scale_x_discrete(labels = c("Strongly Agree", "Agree", "Disagree", "Strongly Disagree", "NA"))
}



```
## Tests
Teen (9th  / 11th grader) confident can do excellent job on 
(fall 2009 / spring 2012) math tests

```{r}

create_grouped_bar_plot(math_eff, "TESTS")


```



## Textbook
Teen (9th  / 11th grader) certain can understand
(fall 2009 / spring 2012) math textbook
```{r}

create_grouped_bar_plot(math_eff, "TEXTBOOK")

```

## Skills
Teen certain can master skills in math course
```{r}

create_grouped_bar_plot(math_eff, "SKILLS")

```
## Assignments 
Teen confident can do an excellent job on math assignments
```{r}

create_grouped_bar_plot(math_eff, "ASSEXCL")

```


## Discrepancy!
From 9th to 11th grade, students' math self-efficacy declines. Why? 

1. Difficulty of coursework goes up, self-efficacy follows as students do poorer
2. 11th graders had more time to compare with their peers than 9th graders did, more comparison = lower self-efficacy?
3. 11th graders have a better gauge of their math ability, are less over-confident. 
4. Higher stakes. 11th graders are thinking about college, where math scores are much more important.
5. Lack of encouragement. 9th graders were highly motivated, just starting high school. 11th graders slack on their assignments / grades, self-efficacy is reflected in this.


Does self-efficacy correlate highly with actual math scores? If yes, could the worst self-efficacy scorers have dropped out? Conveniently, there is a "mathematics ability variable." Let's find out!

## Actual and Efficacy 
Treating the likert scale as continuous for the correlation

```{r}
# Renaming math ability for easier calling

hsls <- dathsls %>%
  rename(math_theta1 = X2TXMTH,
         math_theta2 = X1TXMTH)


summary(hsls$math_theta1)
summary(hsls$math_theta2)

head(math_eff$S1MTESTS)

hist(hsls$math_theta1)
hist(math_eff$S1MTESTS)


cor.test(hsls$math_theta1, hsls$S1MTESTS)

```


```{r}

tests_sagree <- hsls[hsls$S1MTESTS %in% c(1), ]
tests_agree <- hsls[hsls$S1MTESTS %in% c(2), ]
tests_disagree <- hsls[hsls$S1MTESTS %in% c(3), ]
tests_sdisagree <-  hsls[hsls$S1MTESTS %in% c(4), ]

head(tests_sdisagree$S1MTESTS, 20)

print("Strongly disagree")
summary(tests_sdisagree$math_theta1)

print("disagree")
summary(tests_disagree$math_theta1)

print("agree")
summary(tests_agree$math_theta1)

print("Strongly agree")
summary(tests_sagree$math_theta1)

print("Total")
summary(hsls$math_theta1)

```

## Enjoying
```{r}

create_grouped_bar_plot(math_eff, "ENJOYING")
#summary(math_eff$S1MENJOYING)

```

## Enjoying Math and Tests

```{r}

seg4 <- hsls[hsls$S1MENJOYING %in% c(4), ]
seg3 <- hsls[hsls$S1MENJOYING %in% c(3), ]
seg2 <- hsls[hsls$S1MENJOYING %in% c(2), ]
seg1 <- hsls[hsls$S1MENJOYING %in% c(1), ]


create_grouped_bar_plot(seg1, "TESTS")
create_grouped_bar_plot(seg2, "TESTS")
create_grouped_bar_plot(seg3, "TESTS")
create_grouped_bar_plot(seg4, "TESTS")




```

```{r}


mean(is.na(seg1$S2MTESTS))
mean(is.na(seg2$S2MTESTS))
mean(is.na(seg3$S2MTESTS))
mean(is.na(seg4$S2MTESTS))

```
Looking at 9th graders that strongly agree that they are enjoying math, proportion of missing for tests for 11th grade is 12%.
For strongly disagree, 20%

Missing at random? Does this mean anything?


```{r}

print("Assignments")
mean(is.na(seg1$S2MASSEXCL))
mean(is.na(seg2$S2MASSEXCL))
mean(is.na(seg3$S2MASSEXCL))
mean(is.na(seg4$S2MASSEXCL))


print("Skills")
mean(is.na(seg1$S2MTESTS))
mean(is.na(seg2$S2MTESTS))
mean(is.na(seg3$S2MTESTS))
mean(is.na(seg4$S2MTESTS))



mean(is.na(seg1$S2MTESTS))
mean(is.na(seg2$S2MTESTS))
mean(is.na(seg3$S2MTESTS))
mean(is.na(seg4$S2MTESTS))



```










